Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2023-12-11 paper/paper-new.tex
update

\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}

@@

@@

questo per la tassonomia dei diversi modelli ordinali (Tutz, 2022)

@@

\begin{verbatim}
## 
## Call:
## lm(formula = ys ~ x1 + x2, data = dat)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -4.0126 -0.6777 -0.0003  0.6754  4.1015 
## 
## Coefficients:
##              Estimate Std. Error t value Pr(>|t|)    
## (Intercept) -0.003164   0.003161  -1.001    0.317    
## x1           1.005849   0.003146 319.749   <2e-16 ***
## x2           0.498501   0.003156 157.971   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.9996 on 99997 degrees of freedom
## Multiple R-squared:  0.5589, Adjusted R-squared:  0.5589 
## F-statistic: 6.335e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
\end{verbatim}

\begin{verbatim}
## formula: y ~ x1 + x2
## data:    dat
## 
##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
##  probit flexible  1e+05 -122018.81 244049.62 5(0)  9.59e-08 9.1e+00
## 
## Coefficients:
##     x1     x2 
## 1.0064 0.4993 
## 
## Threshold coefficients:
##     1|2     2|3     3|4     4|5 
## -0.8347 -0.2504  0.2529  0.8444
\end{verbatim}

\begin{verbatim}
##  (Intercept)           x1           x2 
## -0.003165429  1.006273800  0.498711694
\end{verbatim}

@@

@@

\begin{verbatim}
##            1         2         3         4         5
## 1 0.19720971 0.2039506 0.2009355 0.1973976 0.2005066
## 2 0.07341871 0.1042636 0.1503074 0.2345890 0.4374213
\end{verbatim}

@@

\begin{verbatim}
##        xa        xb 
## -1.003193 -1.003193
\end{verbatim}

@@

From a statistical point of view, the proportional odds assumption can be assessed by fitting \(k - 1\) binomial regressions and checking if the estimated \(\beta\) is similar between regressions. The regressions are estimated by creating \(k - 1\) dummy variables from the ordinal \(y\). The code below show that simulating data with the proportional odds create \(k - 1\) binomial regression with similar \(\beta\)s.

@@

\begin{verbatim}
## Calls:
## 1: glm(formula = y123vs4 ~ x, family = binomial(link = "logit"), data = dat)
## 2: glm(formula = y12vs34 ~ x, family = binomial(link = "logit"), data = dat)
## 3: glm(formula = y1vs234 ~ x, family = binomial(link = "logit"), data = dat)
## 
##             Model 1 Model 2 Model 3
## (Intercept)  1.1117  0.0148 -1.1014
## SE           0.0151  0.0163  0.0216
##                                    
## x           -5.0263 -5.0329 -4.9763
## SE           0.0365  0.0486  0.0732
## 
\end{verbatim}

@@

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cat\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{m =} \DecValTok{0}\NormalTok{, }
                            \AttributeTok{s =} \DecValTok{1}\NormalTok{, }
                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{),}
                            \AttributeTok{plot =} \FunctionTok{c}\NormalTok{(}\StringTok{"probs"}\NormalTok{, }\StringTok{"latent"}\NormalTok{, }\StringTok{"both"}\NormalTok{))\{}
\NormalTok{  plot }\OtherTok{\textless{}{-}} \FunctionTok{match.arg}\NormalTok{(plot)}
\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
\NormalTok{  x }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"g"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(m))}
\NormalTok{  lat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
    \AttributeTok{x =}\NormalTok{ x,}
    \AttributeTok{m =}\NormalTok{ m,}
    \AttributeTok{s =}\NormalTok{ s}
\NormalTok{  )}

@@

\NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}

@@

\NormalTok{    legend\_b }\OtherTok{\textless{}{-}} \FunctionTok{get\_legend}\NormalTok{(}

@@

\NormalTok{    plts }\OtherTok{\textless{}{-}} \FunctionTok{plot\_grid}\NormalTok{(}

@@

    \FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Cliff2016-ck}
Cliff, N. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Tutz2022-dg}
Tutz, G. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}

>>>>

\newcommand{\CSLBlock}[1]{\hfill\break#1\hfill\break}

@@

\subsection{notation}\label{notation}

\begin{itemize}
\tightlist
\item
  idea di mettere \(F\) o qualcosa di simile per la pdf (\(\Phi\) e.g., per la normale)
\item
  \(Y^\star\) per la latente
\item
  \(k\) numero di livelli ordinali
\item
  \(\alpha\) per le soglie
\end{itemize}

per notazione \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf} e Bürkner and Vuorre (2019)

per la parte di notazione e quella più formale anche DeCarlo (2010)

\subsection{cumulative vs others}\label{cumulative-vs-others}

\begin{itemize}
\tightlist
\item
  cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
\end{itemize}

from Bürkner and Vuorre (2019):

\begin{quote}
For many ordinal variables, the assumption of a single underlying continuous variable, as in cumulative models, may not be appropriate. If the response can be understood as being the result of a sequential pro- cess, such that a higher response category is possible only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
\end{quote}

vedi se fare k-1 logistiche è più o meno la stessa cosa

in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption

come sottolinea Bürkner and Vuorre (2019) e Gerhard Tutz and Berger (2020) i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria \(r\) considerando che la categoria \(r - 1\) è stata raggiunta.

Bürkner and Vuorre (2019) suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages. The probability of passing from one category to the other is modeled as a function of predictors.

@@

questo per la tassonomia dei diversi modelli ordinali (Gerhard Tutz, 2022)

@@

\begin{verbatim}
## 
## Call:
## lm(formula = ys ~ x1 + x2, data = dat)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -4.5222 -0.6765  0.0000  0.6741  4.3425 
## 
## Coefficients:
##             Estimate Std. Error t value Pr(>|t|)    
## (Intercept) 0.001086   0.003159   0.344    0.731    
## x1          0.998374   0.003167 315.209   <2e-16 ***
## x2          0.502929   0.003163 159.004   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.9989 on 99997 degrees of freedom
## Multiple R-squared:  0.5549, Adjusted R-squared:  0.5549 
## F-statistic: 6.233e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
\end{verbatim}

\begin{verbatim}
## formula: y ~ x1 + x2
## data:    dat
## 
##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
##  probit flexible  1e+05 -122644.47 245300.95 5(0)  4.75e-08 9.0e+00
## 
## Coefficients:
##     x1     x2 
## 0.9974 0.5013 
## 
## Threshold coefficients:
##     1|2     2|3     3|4     4|5 
## -0.8473 -0.2529  0.2551  0.8449
\end{verbatim}

\begin{verbatim}
## (Intercept)          x1          x2 
## 0.001086866 0.999502295 0.503497031
\end{verbatim}

@@

The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation). Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion. The main improvement regards mapping the values (for thresholds \(\alpha_i\) and slopes \(\beta\)) from latent standard distribution (gaussian or logistic) into the scale of the \(y\) ordinal value. The scale of the variable depend on the numeric labels assigned to ordered categories. There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to \(\alpha_1 + 0.5\) and \(\alpha_k - 0.5\) where other thresholds are estimated.

Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.

Using the function proposed by Kruschke (2015) (metti ref ad osf) and the equations presented in Kurz (2023) the \texttt{clm\_to\_ord()} function convert parameters fitted with the \texttt{clm} function into the corresponding parameters for the latent \(Y^\star\) variable considering the actual range of values (from 1 to \(k\)).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{clm\_to\_ord }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(fit)\{}
\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(fit}\SpecialCharTok{$}\NormalTok{alpha) }\CommentTok{\# estimated thresholds}
\NormalTok{  coefs }\OtherTok{\textless{}{-}}\NormalTok{ fit}\SpecialCharTok{$}\NormalTok{coefficients}
\NormalTok{  coefs }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(coefs[(}\FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(coefs)])}
\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of levels for y}
\NormalTok{  y }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k }\CommentTok{\# ordinal values}
\NormalTok{  stdm }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# latent mean}
\NormalTok{  stds }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# latent sigma}
\NormalTok{  th\_y }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(th, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{+} \FloatTok{0.5}\NormalTok{, y[k] }\SpecialCharTok{{-}} \FloatTok{0.5}\NormalTok{))}
\NormalTok{  lats }\OtherTok{\textless{}{-}}\NormalTok{ (th\_y[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th\_y[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\SpecialCharTok{/}\NormalTok{ (th[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\CommentTok{\# real latent sigma}
\NormalTok{  b0 }\OtherTok{\textless{}{-}}\NormalTok{ th\_y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ lats}
\NormalTok{  betas }\OtherTok{\textless{}{-}}\NormalTok{ lats }\SpecialCharTok{*}\NormalTok{ coefs}
\NormalTok{  out }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{sigma =}\NormalTok{ lats, }\AttributeTok{b0 =}\NormalTok{ b0, }\AttributeTok{beta =}\NormalTok{ betas, }\AttributeTok{alpha =}\NormalTok{ th\_y)}
  \FunctionTok{return}\NormalTok{(out)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{x =} \FunctionTok{rnorm}\NormalTok{(}\FloatTok{1e5}\NormalTok{)}
\NormalTok{)}

\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
\FunctionTok{clm\_to\_ord}\NormalTok{(fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $sigma
## [1] 1.79423
## 
## $b0
## [1] 2.999998
## 
## $beta
## [1] 1.786346
## 
## $alpha
## [1] 1.500000 2.543349 3.447639 4.500000
\end{verbatim}

\subsection{Gelman parametrization}\label{gelman-parametrization}

rivedi questa parametrizzazione in Gelman, Hill, and Vehtari (2020)

@@

\begin{verbatim}
##            1         2         3         4         5
## 1 0.19881313 0.1989910 0.1974955 0.1998745 0.2048259
## 2 0.07912516 0.1070303 0.1513017 0.2359672 0.4265757
\end{verbatim}

@@

\begin{verbatim}
##         xa         xb 
## -0.9791059 -0.9791059
\end{verbatim}

@@

From a statistical point of view, the proportional odds assumption can be assessed by fitting \(k - 1\) binomial regressions and checking if the estimated \(\beta\) is similar between regressions. The regressions are estimated by creating \(k - 1\) dummy variables from the ordinal \(y\). The code below show that simulating data with the proportional odds create \(k - 1\) binomial regression with similar \(\beta\)s. In fact, fitting \(k - 1\) binomial regressions can be considered also an alternative to fitting an ordinal regression (see Gelman et al., 2020) with more flexibility in parameters (there will be \(k - 1\) regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).

@@

\begin{verbatim}
## Calls:
## 1: glm(formula = y123vs4 ~ x, family = binomial(link = "logit"), data = dat)
## 2: glm(formula = y12vs34 ~ x, family = binomial(link = "logit"), data = dat)
## 3: glm(formula = y1vs234 ~ x, family = binomial(link = "logit"), data = dat)
## 
##               Model 1   Model 2   Model 3
## (Intercept)  1.098911 -0.000607 -1.096559
## SE           0.015143  0.016352  0.021668
##                                          
## x             -5.0004   -4.9958   -5.0132
## SE             0.0364    0.0484    0.0736
## 
\end{verbatim}

spiegare qui la differenza tra proportional and non proportional odds Gerhard Tutz and Berger (2020).

@@

\subsubsection{simulating using probabilities}\label{simulating-using-probabilities}

The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution. This is similar to the general method to simulate data for a generalized linear model where the linear predictors \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component. The following code box simulate a \(n\) binary trials with a continuous predictor \(x\).

{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-9-1}

\caption{ }\label{fig:unnamed-chunk-9}
\end{figure}

We can apply the same idea to an ordinal outcome but we need \(k - 1\) equations where \(k\) is the number of ordinal levels. Let's simulate a similar design with a continuous \(x\) predictor and \(k = 4\) options. We fix the baseline probabilities where \(x = 0\) as uniform thus \(p(y_1) = p(y_2) = ... p(y_k) = 1/k\). The general workflow for the simulation is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
\item
  Define the regression coefficients \(\beta\)s
\item
  Define the baseline probabilities when \(x = 0\). These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
\item
  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\) and regression coefficients \(\mathbf{\beta}\).
\item
  Apply the inverse of the link function \(g^{-1}(\eta)\) on the linear predictor and calculate the cumulative probabilities \(p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)\).
\item
  Calculate for each observation the probability of the \(k\) outcome as the difference between cumulative probabilities. \(p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)\) (sistema qua). This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
\item
  Sample \(n\) outcomes from a multinomial distribution, choosing between \(k\) alternatives with the calculated probabilities.
\end{enumerate}

(rivedi se fare con for)
This can be easily implemented in R. We are using a \texttt{for} loop to improve the clarity of the workflow but using the \texttt{*apply} family of functions can improve the efficiency of the computations.

The linear predictors can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).

{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-11-1}

\caption{ }\label{fig:unnamed-chunk-11}
\end{figure}

Now we can apply the inverse of the link function \(g^{-1} = \Phi^{-1}\) to calculate the corresponding cumulative probabilities.

\begin{verbatim}
##    p_y1  p_y2  p_y3  p_y4
## 1  0.57 0.232 0.134 0.063
## 2 0.128 0.194 0.262 0.415
## 3   0.1 0.172 0.255 0.472
## 4 0.043 0.105 0.207 0.645
## 5   ...   ...   ...   ...
## 6 0.381 0.264 0.207 0.147
## 7 0.459 0.257 0.177 0.106
## 8 0.332 0.263 0.225  0.18
## 9 0.439  0.26 0.185 0.116
\end{verbatim}

We can plot the expected effect of \(\beta_1\) on the \(k\) probabilities using the \texttt{num\_latent\_plot()} function (see Figure \ref{fig:sim-with-probs-lat-plot}). Finally we sample using the \texttt{sample()} function using the calculated probabilities.

{\centering \includegraphics{paper-new_files/figure-latex/sim-with-probs-lat-plot-1}

\caption{sim-with-probs-lat-plot}\label{fig:sim-with-probs-lat-plot}
\end{figure}

To check the simulation result we can increase the number of observations, fit the model using the \texttt{ordinal::clm()} function and assess the recovery of simulated parameters.

\subsubsection{simulating using latent distribution}\label{simulating-using-latent-distribution}

An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model. This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and the cutting the latent values according to the thresholds \(\alpha\). The workflow is slighlty different compared to the previous paragraph:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
\item
  Define the regression coefficients \(\beta\)
\item
  Define the baseline probabilities when \(x = 0\). These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
\item
  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\), the regression coefficients \(\mathbf{\beta}\) and the error \(\mathbf{\epsilon}\) sampled from the appropriate probability distribution.
\item
  Cut the latent variable into \(k\) areas using the thresholds \(\alpha\) and assign the corresponding ordinal value. This step simply checks if the latent value \(Y^{\star}_i\) is between two threshold values and assign the corresponding value. This can be done using the \texttt{cut()} or the \texttt{findInterval()} functions.
\end{enumerate}

The Figure (put crossref) depict the simulated \(Y^{*}\) and the corresponding ordinal value. As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.

{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-18-1}

\caption{ }\label{fig:unnamed-chunk-18}
\end{figure}

The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sim\_ord\_latent }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(location,}
                           \AttributeTok{scale =} \ConstantTok{NULL}\NormalTok{,}
\NormalTok{                           By,}
                           \AttributeTok{Bscale =} \ConstantTok{NULL}\NormalTok{,}
                           \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
                           \AttributeTok{alphas =} \ConstantTok{NULL}\NormalTok{, }
\NormalTok{                           data, }
                           \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{))\{}

  \CommentTok{\# get all the correct functions according to the link}
\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(alphas))\{}
    \CommentTok{\# calculate thresholds if not provided}
\NormalTok{    alphas }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, }\AttributeTok{link =}\NormalTok{ link)}
\NormalTok{  \}}
\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of ordinal outcomes}
\NormalTok{  n }\OtherTok{\textless{}{-}} \FunctionTok{nrow}\NormalTok{(data) }\CommentTok{\# number of observations}

  \CommentTok{\# model matrix for the location effect}
\NormalTok{  Xy }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(location, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
\NormalTok{  lpy }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xy }\SpecialCharTok{\%*\%}\NormalTok{ By) }\CommentTok{\# linear predictor for the location}
\NormalTok{  lps }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# default scale effect 0, exp(0) = 1 (the default scale)}

  \CommentTok{\# check predictors on the scale parameter}
  \ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!}\FunctionTok{is.null}\NormalTok{(scale))\{}
    \CommentTok{\# model matrix for the scale effect}
\NormalTok{    Xsigma }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(scale, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
\NormalTok{    lps }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xsigma }\SpecialCharTok{\%*\%}\NormalTok{ Bscale) }\CommentTok{\# linear predictor for the scale}
\NormalTok{  \}}

  \CommentTok{\# latent variable with appropriate error function}
\NormalTok{  ystar }\OtherTok{\textless{}{-}}\NormalTok{ lf}\SpecialCharTok{$}\FunctionTok{rfun}\NormalTok{(n, lpy, }\FunctionTok{exp}\NormalTok{(lps))}

  \CommentTok{\# cut according to thresholds}
\NormalTok{  y }\OtherTok{\textless{}{-}} \FunctionTok{findInterval}\NormalTok{(ystar, alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# to start from 1}

\NormalTok{  data}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{ordered}\NormalTok{(y) }\CommentTok{\# to ordered factor}
\NormalTok{  data}\SpecialCharTok{$}\NormalTok{ys }\OtherTok{\textless{}{-}}\NormalTok{ ystar }\CommentTok{\# save also the latent}
  \FunctionTok{return}\NormalTok{(data)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori Gerhard Tutz and Berger (2020)

dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds Christensen (2019)

se citi brms per modelli più complicati, ricordati di specificare che:

\begin{quote}
That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
\end{quote}

un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale. La slope della funzione cumulative corrisponde alla varianza. Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in \texttt{brms} si chiama desc)

si può citare anche questo Cox (1995) come riferimento per i modelli

una distinzione tra location-shift and location-scale models G. Tutz and Berger (2017) dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.

sempre in G. Tutz and Berger (2017) propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds

\subsubsection{Simulating scale effects}\label{simulating-scale-effects}

The location-scale model can be simulated using the \texttt{sim\_ord\_latent()} function and providing the predictors for the scale parameters. Given the \texttt{log} link function, predictors are provided on the \texttt{log} scale. For example, we simulate the effect of a binary variable \(x\) representing two independent groups predicting the \(k = 5\) response. We simulate a \emph{location} effect of \(\beta_1 = 0.5\) (in \emph{probit} scale) and \(\zeta_1 = \text{log}(2) = 0.70\). The first group has a \(\sigma = 1\) and the second group has \(\sigma = 2\). Again we simulate that the baseline probabilities are uniform for the first group.

The table (put ref) reports the simulation results.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cat\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{m =} \DecValTok{0}\NormalTok{, }
                            \AttributeTok{s =} \DecValTok{1}\NormalTok{, }
                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{),}
                            \AttributeTok{plot =} \FunctionTok{c}\NormalTok{(}\StringTok{"probs"}\NormalTok{, }\StringTok{"latent"}\NormalTok{, }\StringTok{"both"}\NormalTok{))\{}
  \FunctionTok{require}\NormalTok{(ggplot2)}
\NormalTok{  plot }\OtherTok{\textless{}{-}} \FunctionTok{match.arg}\NormalTok{(plot)}
\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
\NormalTok{  x }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"g"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(m))}
\NormalTok{  lat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
    \AttributeTok{x =}\NormalTok{ x,}
    \AttributeTok{m =}\NormalTok{ m,}
    \AttributeTok{s =}\NormalTok{ s}
\NormalTok{  )}

@@

\NormalTok{  latl }\OtherTok{\textless{}{-}}\NormalTok{ tidyr}\SpecialCharTok{::}\FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}

@@

\NormalTok{    legend\_b }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{get\_legend}\NormalTok{(}

@@

\NormalTok{    plts }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(}

@@

\NormalTok{    cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Burkner2019-aw}
Bürkner, P.-C., \& Vuorre, M. (2019). Ordinal regression models in psychology: A tutorial. \emph{Advances in Methods and Practices in Psychological Science}, \emph{2}, 77--101. \url{https://doi.org/10.1177/2515245918823199}

\bibitem[\citeproctext]{ref-Christensen2019-cz}
Christensen, R. H. B. (2019). \emph{Ordinal---regression models for ordinal data}.

\bibitem[\citeproctext]{ref-Cliff2016-ck}
Cliff, N. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.

\bibitem[\citeproctext]{ref-Cox1995-ur}
Cox, C. (1995). Location-scale cumulative odds models for ordinal data: A generalized non-linear model approach. \emph{Statistics in Medicine}, \emph{14}, 1191--1203. \url{https://doi.org/10.1002/sim.4780141105}

@@

\bibitem[\citeproctext]{ref-DeCarlo2010-lj}
DeCarlo, L. T. (2010). On the statistical and theoretical basis of signal detection theory and extensions: Unequal variance, random coefficient, and mixture models. \emph{Journal of Mathematical Psychology}, \emph{54}, 304--313. \url{https://doi.org/10.1016/j.jmp.2010.01.001}

@@

\bibitem[\citeproctext]{ref-Gelman2020-tg}
Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and other stories}. Cambridge University Press. \url{https://doi.org/10.1017/9781139161879}

@@

\bibitem[\citeproctext]{ref-Kurz2023-zv}
Kurz, A. S. (2023). \emph{Doing bayesian data analysis in brms and the tidyverse} (Version 1.1.0). Retrieved from \url{https://bookdown.org/content/3686/}

@@

\bibitem[\citeproctext]{ref-Tutz2022-dg}
Tutz, Gerhard. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}

\bibitem[\citeproctext]{ref-Tutz2017-de}
Tutz, G., \& Berger, M. (2017). Separating location and dispersion in ordinal regression models. \emph{Econometrics and Statistics}, \emph{2}, 131--148. \url{https://doi.org/10.1016/j.ecosta.2016.10.002}

\bibitem[\citeproctext]{ref-Tutz2020-xq}
Tutz, Gerhard, \& Berger, M. (2020). Non proportional odds models are widely dispensable -- sparser modeling based on parametric and additive location-shift approaches. Retrieved from \url{http://arxiv.org/abs/2006.03914}
